NumLin: Linear Types for Linear Algebra

### 2018-11-16 write-up/paper/discussion.tex
Finish Related Work

% The main point we want to make is that using linear types for BLAS
% is an ``obvious'' idea, but is surprisingly under-explored.

@@

ATS\cite{cui2005ats} is a language which combines dependent and linear types;
although it provides BLAS bindings, the main focus in the API provided there is
for statically enforcing dimensions related constraints of the arguments,
rather than aliasing restrictions as shown here.

\begin{itemize}
\item Single-assignment C: for hardware, doesn't allow higher-order fun.
\item Linear Haskell: linearity on the arrows.
\item Bernardy and Sveningsson: very theoretical, no implementation.
\item L3: can use existential types to statically track pointer identities to
    prevent that source of runtime errors.
\item Boyland fractional permissions
\item Disadvantage of graph-based approaches to matrix expression generation:
    buggy, like SymPy.
\item multi-stage: once dimensions known, they are fixed, so we can take
    dimension checking one-step further and implement a matrix chaining alg.
\end{itemize}

% Dimensions: easy enough to integrate, even w/o ATS/dep types
% generative phantom types
% \cite{abe2015simple}

@@

We are pleasantly surprised at how simple the overall design and implementation
of \lang\ is, given its expressive power and usability. Indeed, the focus on
getting a working prototype early on (so that we could test it with real
BLAS/LAPACK routines as soon as possible) meant that we only added features to
the type system when it was clear that they were absolutely necessary: these
features were !-types and value-restriction for the \highl{Many} constructor.
Although we considered parametric polymorphism, size-types and tracking pointer
identities at the type-level\cite{morrisett} (all of which would certainly be
useful and are potential avenues for further work) it turns out these features
are not necessary.

We believe we have stumbled across a particularly nice problem-solution fit:
the idea of linear types for linear algebra -- one that used to just be
commonly believed folklore -- actually has some merit and is worth exploring.
We also think that this concept (and the general design of its implementation)
need not be limited to linear algebra: we could conceivably `backport' this
idea to other contexts that need linearity (concurrency, single-use
continuations, zero-copy buffer, streaming I/O) or combine it with dependent
types to achieve even more expressive power to split up a single block of
memory into multiple regions in an arbitrary manner\cite{space_monads}.

>>>>

@@

Linear Haskell\cite{bernardy2017linear}, takes a slightly different definition
of linearity, that is one on \emph{arrows} as opposed to \emph{kinds}: for $f :
a \multimap b$, if $f u$ is used exactly once \emph{then} $u$ is used exactly
once. Whilst this has the advantage of being backwards-compatible, \lang's
approach of compile-time code-generation enables using linearity \emph{as a
library}, whilst the use of fractions gives extra control over aliasing not
available in Linear Haskell.

In general, using substructural types to express array computations is not
particularly new\cite{scholz,henriksen,bernardy2016}, but these approaches
limit recursion and higher-order functions, and have special constructs to
enable \emph{implementation} of efficient BLAS-like libraries rather than
enforcing correct \emph{usage}.

@@

We are pleasantly surprised at how simple the overall design and implementation
of \lang\ is, given its expressive power and usability.  So simple in fact,
that fractions, a convenient theoretical abstraction until this point, could be
implemented by restricting division and multiplication to be by 2 only
\cite{boyland2003}, thus turning any required arithmetic into unification.

Indeed, the focus on getting a working prototype early on (so that we could
test it with real BLAS/LAPACK routines as soon as possible) meant that we only
added features to the type system when it was clear that they were absolutely
necessary: these features were !-types and value-restriction for the
\highl{Many} constructor.

Going forwards, one may wish to eliminate even more runtime errors from \lang,
by extending its type system. For example, we could have used existential types
to statically track pointer identities\cite{morrisett}, or parametric
polymorphism.

We could also attempt to catch mismatched dimensions at compile time as well.
While this could be done with generative phantom types\cite{abe2015simple},
using dependent types may offer more flexibility in \emph{partitioning}
regions\cite{space_monads} or statically enforcing dimensions related
constraints of the arguments at compile-time.  ATS\cite{cui2005ats} is just
such a language which combines dependent and linear types; although it provides
BLAS bindings, it does not aim to provide aliasing restrictions as demonstrated
in this paper.

Taking this idea one step even further, since matrix dimensions are typically
fixed at runtime, we could \emph{stage} \lang\ programs and compile matrix
expressions using more sophisticated algorithms\cite{barthels}. However, it is
worth noting that without care, such algorithms\cite{rocklin_thesis}, usually
based on graph-based, ad-hoc dataflow analysis, can produce erroneous output
which would not get past a linear type system with fractions.

We also think that this concept (and the general design of its implementation)
need not be limited to linear algebra: we could conceivably `backport' this
idea to other contexts that need linearity (concurrency, single-use
continuations, zero-copy buffer, streaming I/O) or combine it with dependent
types to achieve even more expressive power to split up a single block of
memory into multiple regions in an arbitrary manner\cite{space_monads}.

### 2018-11-16 write-up/paper/lang_and_examples.tex
Start Performance Metrics
Finish Related Work

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
    \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the
        equations this code implements. Line numbers in comments refer to
        equivalent lines in a C implementation
        (Figure~\ref{fig:cblas_kalman}).}\label{fig:lang_kalman}
\end{figure}

@@

    \item \texttt{potrs \_ chol\_r data\_2} uses a pre-computed Cholesky decomposition
        to multiply \texttt{data\_2} $ = H \mu - \textrm{data}$ by
        \texttt{r\_2} $= (R + H \Sigma H^T)^{-1}$

    \item a lot of memory re-use; the following sets of identifiers alias each other:
        \begin{itemize}
            \item \texttt{r\_1} and \texttt{r\_2}
            \item \texttt{data\_1}, \texttt{data\_2} and \texttt{sol\_data}
            \item \texttt{new\_h} and \texttt{sol\_h}
            \item \texttt{h\_sol\_h}, \texttt{sigma\_copy} and \texttt{new\_sigma}
            \item \texttt{mu\_copy} and \texttt{new\_mu}.
        \end{itemize}

>>>>

\begin{figure}[t]
    \centering
%   \begin{minipage}{.4\textwidth}
%   \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
%   \end{minipage}
%   \begin{minipage}{.4\textwidth}
%   \centering
%   \begin{minted}[linenos,fontsize=\small]{ocaml}
%   let owl_kalman
%       ~sigma
%       ~h
%       ~mu
%       ~r
%       ~data =
%     let open Owl.Mat in
%     let ( * ) = dot in
%     let h' = transpose h in
%     let sigma_h' = sigma * h' in
%     let x = sigma_h' * (inv @@ r + h * sigma_h') in
%     let new_mu = mu + x * (h * mu - data) in
%     let new_sigma = sigma - x * h * sigma in
%     new_sigma, new_mu
%   ;;
%   \end{minted}
%   \end{minipage}
    \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the
        equations this code implements. Line numbers in comments refer to
        equivalent lines in a C implementation
        (Figure~\ref{fig:cblas_kalman}).}\label{fig:lang_kalman}
\end{figure}

\begin{figure}[t]
    \centering
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{grammar}
            <f> ::= `'
            \alt <fc>
            \alt `Z'
            \alt `S' <f>

            <t> ::= `'
            \alt `unit'
            \alt `bool'
            \alt `int'
            \alt `elt'
            \alt <f> `arr'
            \alt <f> `mat'
            \alt `!' <t>
            \alt \synt{$\forall$} <fc.> <t>
            \alt <t> \lit{$\otimes$} \synt{$t'$}
            \alt <t> \lit{$\multimap$} \synt{$t'$}
        \end{grammar}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{minted}[fontsize=\small]{ocaml}
module Arr =
  Owl.Dense.Ndarray.D

type z = Z
type 'a s = Succ

type 'a arr =
  A of Arr.arr
  [@@unboxed]

type 'a mat =
  M of Arr.arr
  [@@unboxed]

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{\lang's type grammar (left) and its embedding into OCaml
        (right).}\label{fig:type_grammar}
\end{figure}

@@

    \item \texttt{posvFlip} is like \texttt{posv} except for solving $XA = B$

    \item a lot of memory re-use; the following sets of identifiers alias each other:
        \begin{itemize}
            \item \texttt{r\_1}, \texttt{r\_2} and \texttt{k\_by\_k}
            \item \texttt{data\_1} and \texttt{data\_2}
            \item \texttt{mu} and \texttt{new\_mu}
            \item \texttt{sigma\_hT} and \texttt{x}
        \end{itemize}
